Fast approximate Bayesian inference of HIV indicators using PCA adaptive Gauss-Hermite quadrature

### 2023-06-28 src/docs_paper/paper.Rmd
Edit to abstract

abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of policy interest, including HIV prevalence, HIV incidence, and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data. The model is provided as a tool for countries to input their data to and generate estimates during a yearly process supported by UNAIDS. Inference has previously been conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package. We propose a new inference method extending adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters, enabling fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method improves the accuracy of inferences across a range of metrics, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler. Our implementation uses the \texttt{aghq} \textsc{R} package, facilitating easy, flexible use of the method when provided a \texttt{TMB} \textsf{C++} template for the model's log-posterior.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - Bayesian statistics
  - spatial statistics
  - evidence synthesis
  - small-area estimation
  - approximate inference
  - INLA
  - AGHQ
  - HIV epidemiology

>>>>

abstract: |
  | Naomi is a spatial evidence synthesis model used to produce district-level HIV epidemic indicators in sub-Saharan Africa. Multiple outcomes of policy interest, including HIV prevalence, HIV incidence, and antiretroviral therapy treatment coverage are jointly modelled using both household survey data and routinely reported health system data. The model is provided as a tool for countries to input their data to and generate estimates during a yearly process supported by UNAIDS. Inference has previously been conducted using empirical Bayes and a Gaussian approximation via the \texttt{TMB} \textsc{R} package. We propose a new inference method extending adaptive Gauss-Hermite quadrature to deal with >20 hyperparameters, enabling fast and accurate inference for Naomi and other extended latent Gaussian models. Using data from Malawi, our method improves the accuracy of inferences across a range of metrics, while being substantially faster to run than Hamiltonian Monte Carlo with the No-U-Turn sampler. Our implementation is based on the existing \texttt{TMB} \textsf{C++} template for the model's log-posterior, and is compatible with any model with such a template.
keyword-subclass: | 
 \begin{keyword}[class=MSC2020] % It must be define for aap, aop, aos journals. For aoas, sts is not used
 \kwd[Primary ]{00X00}
 \kwd{00X00}
 \kwd[; secondary ]{00X00}
 \end{keyword}
keywords: 
  - Bayesian statistics
  - spatial statistics
  - evidence synthesis
  - small-area estimation
  - approximate inference
  - INLA
  - AGHQ
  - HIV epidemiology
